---
title: 'Autoencoder and Deep Features'
author: 'Loic Merckel'
date: '22 July 2017'
output:
  html_document:
    number_sections: false
    toc: true
    highlight: tango
    theme: cosmo
    smart: true
---

<style type="text/css">
  h1.title { font-weight: bold; } h1 { font-weight: normal; } .author { font-weight: normal; font-size: 1.5em; }
</style>


```{r include=FALSE}
# License -----------------

# Copyright 2017 Loic Merckel
# 
# Licensed under the Apache License, Version 2.0 (the "License");
# you may not use this file except in compliance with the License.
# You may obtain a copy of the License at
# 
# http://www.apache.org/licenses/LICENSE-2.0
# 
# Unless required by applicable law or agreed to in writing, software
# distributed under the License is distributed on an "AS IS" BASIS,
# WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
# See the License for the specific language governing permissions and
# limitations under the License.
```

We use the autoencoder and deeplearning capability of the  [H<small>2</small>O](https://www.h2o.ai/) framework to explore deep features. 

Such an approach provides a means of reducing features (although here there is no need, for the number of features is already small). 

It also offers a way to leverage unlabelled data---a technique called semi-supervised learning. The kernels *[Diabolo Trick: Pre-train, Train and Solve](https://goo.gl/JXxc6n)* gives a concrete example.

```{r include=FALSE}
pkgs <- c("corrplot", "caret", "data.table", "plyr",
          "xgboost", "parallel", "Metrics", "maps", 
          "ggmap", "lubridate", "fasttime", "gridExtra") 
for (pkg in pkgs) {
  if (! (pkg %in% rownames(installed.packages()))) { install.packages(pkg) }
  require(pkg, character.only = TRUE)
}
rm(pkgs)
rm(pkg)
```

```{r include=FALSE}
# data ---------------------------
rm(list=ls(all=TRUE))

kIsOnKaggle <- FALSE
if (kIsOnKaggle) {
  X <- fread(file.path("..", "input", "train.csv", fsep = .Platform$file.sep), 
             header = TRUE, data.table = TRUE, na.strings=c("NA","?", ""))
} else {
  X <- fread(file.path(".", "data", "train.csv", fsep = .Platform$file.sep), 
             header = TRUE, data.table = TRUE, na.strings=c("NA","?", ""))
}
```


```{r include=FALSE}
isConstant <- function (x) {
  return(all(duplicated(x)[-1L]))
}
```

# Features Tinkering

First, we remove the column `dropoff_datetime`, as otherwise there is no point to predict the duration of the trip... We could just precisely calculate it...

```{r  echo=TRUE, warning=FALSE, results='hide', message=FALSE}
X[, dropoff_datetime := NULL]
```

```{r include=FALSE}
toIgnore <- c("id", "trip_duration")
```


## Outliers

### Pickup Location

```{r echo=FALSE, warning=FALSE, results='hide', message=FALSE, out.width ="60%"}
ggmapSupported <- TRUE
tryCatch({
map <- get_map(
  location = c(lon = median(X$pickup_longitude), 
               lat = 40.65), #median(X$pickup_latitude)
  zoom = 11)

outliers <- which(X$pickup_longitude > -74
                  & X$pickup_latitude < 40.57)

set.seed(1)
nonOutlierSample <- sample(setdiff(seq(1, nrow(X)), outliers), 1000)

ggmap(map) + 
  geom_point(data = X[outliers, ],
             aes(x = pickup_longitude, 
                 y = pickup_latitude), color = "red", size = 4) + 
  geom_point(data = X[nonOutlierSample, ],
             aes(x = pickup_longitude, 
                 y = pickup_latitude), color = "blue", size = 4, shape = 1)
}, error = function(e) {
  print (e)
  ggmapSupported <<- FALSE
})
```

`r if(!ggmapSupported) {"<img src=\"https://raw.githubusercontent.com/roikku/kaggle/master/taxi/img/map.png\" width=\"60%\"/>"}`

```{r include=FALSE}
rm(ggmapSupported)
```

There seems to be some outliers in the dataset. Let's remove *some* of them (obviously, a better way should be devised here, for the current naive implementation misses some outliers while removing some non-outliers).

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
X <- X[-which(X$pickup_longitude > -74
             & X$pickup_latitude < 40.57), ]
```

```{r include=FALSE}
rm(map, outliers, nonOutlierSample)
```

### Trip Duration

```{r include=FALSE}
showOutliers <- function (X, column, maxVal, histTitle) {
  # with outlier
  opar <- par(fig=c(0, 0.5, 0, 0.8))
  hist(x = X[[column]], breaks = 100, col = "forestgreen", plot = TRUE, xlab = column, 
       main = NULL)
  par(fig=c(0, 0.5, 0.50, 1), new = TRUE)
  boxplot(X[[column]], col = "forestgreen", outcol="firebrick4", horizontal = TRUE, axes = FALSE)
  mtext("With outliers", outer = FALSE, line=0)
  
  # without outlier
  par(fig=c(0.5, 1, 0, 0.8), new = TRUE)
  hist(x = X[-which(X[[column]] > maxVal), ][[column]], breaks = 100, col = "forestgreen", plot = TRUE, xlab = column, 
       main = NULL)
  par(fig=c(0.5, 1, 0.50, 1), new = TRUE)
  boxplot(X[-which(X[[column]] > maxVal), ][[column]], col = "forestgreen", outcol="firebrick4", horizontal = TRUE, axes = FALSE)
  mtext("Without outliers", outer = FALSE, line=0)
  title(histTitle, outer = TRUE, line=-2)
  par (opar)
  return (0)
}
```

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE, fig.width=10, fig.height=5}
# outliers (Hampel's test)
maxThreshold <- 3 * 1.48 * mad(X$trip_duration, na.rm = TRUE) 
                    + median(X$trip_duration, na.rm = TRUE)  
showOutliers(X, "trip_duration", maxThreshold, "Distribution of Trip Duration")
```

We rely on the Hampel's test to compute the maximum threshold above which values are declared spurious. One can find further details about the Hampel's test for outlier detection in *[Scrub data with scale-invariant nonlinear digital filters](http://m.eet.com/media/1140823/191159.pdf)*.

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
X <- X[-which(X$trip_duration > maxThreshold), ]
```

```{r include=FALSE}
rm(maxThreshold)
```

## Pickup Date

We add the two new features `wday` and `hour`.

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
X[, pickup_datetime := fastPOSIXct (pickup_datetime)]
X[, wday := wday(pickup_datetime)]
X[, hour := hour(pickup_datetime)]
```

```{r include=FALSE}
toIgnore <- c(toIgnore, c("pickup_datetime"))
```

## Convert Categorical Variables to Numerical

```{r include=FALSE}
numVal <- c()
nonNumVal <- c()
for (feature in colnames(X)) {
  if (feature %in% toIgnore) {
    next
  }
  if (class(X[[feature]]) %in% c("integer", "numeric")) {
    if (sd(X[[feature]]) == 0) {
      toIgnore <- c(toIgnore, feature)
    } else {
      numVal <- c(feature, numVal)
    }
  } else {
    nonNumVal <- c(feature, nonNumVal)
  }
}
rm(feature)
```

There is `r length(nonNumVal)` categorical variables, and there are `r length(numVal)` numerical variables with non-null standard deviation. 

```{r echo=TRUE, warning=FALSE, message=FALSE}
X[, store_and_fwd_flag := ifelse (store_and_fwd_flag == "N", 0, 1)]
```


# Deep Features

## Convert to H2O Frames

```{r include=FALSE}
if (! ("h2o" %in% rownames(installed.packages()))) { install.packages("h2o") }
require("h2o")

tryCatch(
  if (h2o.clusterIsUp()) {
    h2o.shutdown(prompt=FALSE)
    Sys.sleep(5)
  }, error = function(e) {
    
  })

h2o.init(nthreads = parallel:::detectCores(), 
         max_mem_size = "15g", min_mem_size = "1g")
h2o.removeAll()
h2o.clusterStatus()
```

```{r include=FALSE}
# We remove the column pickup_datetime because H2O does not handle POSIXct type, 
# and we do not really need this column any longer (wday and hour are used instead)

X[, pickup_datetime := NULL]
```

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}

if (kIsOnKaggle) {
  # we reduce the training set size to avoid time out
  set.seed(1)
  trainIndexes <- createDataPartition(
    y = X[["trip_duration"]], p = 0.1, list = FALSE) 
  X <- X[trainIndexes, ]
  rm(trainIndexes)
}

train <- as.h2o(X)
```

```{r echo=TRUE, warning=FALSE, message=FALSE}
predictors <- setdiff (names(X), c(toIgnore))
```

## Autoencoder (Diabolo Network)

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
if (kIsOnKaggle) {
  hyperParamsAutoencoder = list( 
    hidden = list(c(8, 4, 8), c(8, 3, 8), c(7, 3, 7), c(6, 3, 6)),
    activation = c("Tanh"))
} else {
  hyperParamsAutoencoder = list( 
    hidden = list(c(11, 8, 11), c(10, 8, 10), c(9, 5, 9), c(8, 5, 8), 
                  c(7, 5, 7), c(6, 5, 6), c(11, 7, 11), c(10, 7, 10), 
                  c(9, 4, 9), c(8, 4, 8), c(7, 4, 7), c(6, 4, 6), 
                  c(11, 6, 11), c(10, 6, 10), c(9, 3, 9), c(8, 3, 8), 
                  c(7, 3, 7), c(6, 3, 6), c(11, 5, 11), c(10, 5, 10),
                  c(11, 4, 11), c(10, 4, 10), c(11, 8, 5, 8, 11)),
    activation = c("Tanh"))
}

gridAutoencoder <- h2o.grid(
  x = predictors,
  autoencoder = TRUE,
  training_frame = train,
  hyper_params = hyperParamsAutoencoder,
  search_criteria = list(strategy = "Cartesian"),
  algorithm = "deeplearning",
  grid_id = "grid_autoencoder", 
  reproducible = TRUE, 
  seed = 1,
  variable_importances = TRUE,
  categorical_encoding = "AUTO",
  score_interval = 2,
  epochs = 800,
  adaptive_rate = TRUE,
  standardize = TRUE,
  ignore_const_cols = FALSE)
```

```{r include=FALSE}
rm (hyperParamsAutoencoder)
```

The following table summarizes the grid results (it is sorted increasingly by 'mse'):
```{r echo=FALSE}
sortedGridAutoencoder <- h2o.getGrid("grid_autoencoder", 
                                     sort_by = "mse", decreasing = FALSE)
tmpDf <- as.data.frame(sortedGridAutoencoder@summary_table)
knitr::kable(head(tmpDf[, -grep("model_ids", colnames(tmpDf))]), row.names = TRUE)
rm(tmpDf)
```

```{r include=FALSE}
bestAutoencoder <- h2o.getModel(sortedGridAutoencoder@model_ids[[1]])

bestAutoencoderErr <- as.data.frame(h2o.anomaly(bestAutoencoder, 
                                                train, 
                                                per_feature = FALSE))
```

Considering the "best" autoencoder (i.e., the one with the lowest 'mse', which is the one with the hidden layers [`r bestAutoencoder@parameters$hidden`]), the two following figures illustrate the fact that it performs rather well; only a limited portion of the input signal could not be reconstructed. 

```{r  echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.width=10, fig.height=3.5}
plotReconstructionError <- function (error) {
  cut <- 0.5 * sd(error)
  sortedErr <- sort(error)
  sortedErrFrame <- data.frame (index = seq(0, length(sortedErr)-1), error = sortedErr)
  ylim <- c(min(sortedErrFrame$error), max(sortedErrFrame$error))
  xlim <- c(min(sortedErrFrame$index), max(sortedErrFrame$index))
  # could do as in https://stackoverflow.com/questions/11838278/plot-with-conditional-colors-based-on-values-in-r
  plot (x = sortedErrFrame$index[which(sortedErrFrame$error <= cut)], 
        y = sortedErrFrame$error[which(sortedErrFrame$error <= cut)], 
        type = "o", col="forestgreen", lwd=1, ylim = ylim, xlim = xlim, ylab="mse",
        main = "Reconstruction Error",
        xlab = "Sorted Index")
  par(new=TRUE)
  plot (x = sortedErrFrame$index[which(sortedErrFrame$error > cut)], 
        y = sortedErrFrame$error[which(sortedErrFrame$error > cut)],  
        type = "o", col="firebrick3", xaxt='n', yaxt='n', ann=FALSE, ylim = ylim, xlim = xlim, xlab="")
  return (0)
}

layout(matrix(c(1,2,2), 1, 3, byrow = TRUE))
plotReconstructionError (bestAutoencoderErr$Reconstruction.MSE)

# https://stackoverflow.com/questions/21858394/partially-color-histogram-in-r
h <- hist(x = bestAutoencoderErr$Reconstruction.MSE, breaks = 100, plot = FALSE)
cuts <- cut(h$breaks, c(-Inf, 0.5 * sd(bestAutoencoderErr$Reconstruction.MSE), Inf))
plot(h, col = c("forestgreen","firebrick3")[cuts], main = "Reconstruction Error", xlab = "mse", lty="blank")
rm(h, cuts)
```


## Deep Features Visualization

```{r  include=FALSE}
plotDeepFeatures <- function(data, maxPlot = 16, ncol = 4) {
  count <- 1
  plotList <- list()
  n <- (ncol(data) - 1)
  midpoint <- median(data$log_trip_duration)
  for (i in 1:n) {
    for (j in 1:n) {
      if (i == j) {
        next
      }
      plotList[[paste0("p", count)]] <- ggplot(
        data, 
        aes_string(
          x = paste0("DF.L", layer, ".C", i), 
          y = paste0("DF.L", layer, ".C", j), 
          color = "log_trip_duration")) +
        geom_point(alpha = 0.9, aes(colour = log_trip_duration)) +
        #scale_colour_gradient2(midpoint = midpoint) + 
        #scale_colour_gradientn(colours = terrain.colors(10)) +
        #scale_colour_gradientn(colours = topo.colors(10)) +
        #scale_colour_gradientn(colours = rev(rainbow(10))) +
        scale_colour_gradientn(colours = rev(rainbow(10)[1:8])) +
        theme(legend.position = 
                ifelse(count == min((n-1)*n, maxPlot), "right", "none")) +     
        labs(color="log(t)")
      
      count <- count + 1
      if (count > maxPlot) {
        break
      }
    }
    if (count > maxPlot) {
      break
    }
  }
  grid.arrange(grobs = as.list(plotList), ncol = ncol)
}
```


### Second Layer

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
layer <- 2
```

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
deepFeature2 <- h2o.deepfeatures(bestAutoencoder, train, layer = layer)
```

```{r echo=TRUE, warning=FALSE, results='show', message=FALSE}
data <- as.data.frame(deepFeature2)
data$log_trip_duration <- log(X$trip_duration)

summary(data)
```

```{r  echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.width=10, fig.height=5}
plotDeepFeatures(data, 8)
```
  
### Third Layer

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
layer <- 3
```

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
deepFeature3 <- h2o.deepfeatures(bestAutoencoder, train, layer = layer)
```

```{r echo=TRUE, warning=FALSE, results='show', message=FALSE}
data <- as.data.frame(deepFeature3)
data$log_trip_duration <- log(X$trip_duration)

summary(data)
```

```{r  echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.width=10, fig.height=10}
plotDeepFeatures(data, 16)
```